quollr: An R Package for Visualizing 2-D Models from Non-linear Dimension Reductions in High Dimensional Space

### 2025-08-04 paper-quollr.Rmd
Fixed some issue with new data

scurve_model_obj <- fit_highd_model(
  highd_data = scurve, 
  nldr_data = scurve_umap, 
  b1 = 15, 
  q = 0.1, 
  benchmark_highdens = 5)

@@

The paper is organized as follows. The next section introduces the implementation of the `quollr` package on CRAN, including a demonstration of the package's key functions and visualisation capabilities. In the application section, we illustrate the algorithm's functionality for studying a clustering data structure. Finally, we conclude the paper with a brief summary and discuss potential opportunities for using our algorithm.

@@

Our algorithm includes the following steps: (1) scaling the NLDR data, (2) computing configurations of a hexagon grid, (3) binning the data, (4) obtaining the centroids of each bin, (5) indicating neighboring bins with line segments that connect the centroids, and (6) lifting the model into high dimensions (`r knitr::asis_output(if (knitr::is_html_output()) { "Figure \\@ref(fig:algo-step-html)"} else { "Figure \\@ref(fig:algo-steps)"})`). A detailed description of the algorithm can be found in **cite the methodology paper**.

@@

`scurve` is a $7\text{-}D$ simulated dataset. It is constructed by simulating $5000$ observations from $\theta \sim U(-3\pi/2, 3\pi/2)$, $X_1 = \sin(\theta)$, $X_2 \sim U(0, 2)$ (adding thickness to the S), $X_3 = \text{sign}(\theta) \times (\cos(\theta) - 1)$. The remaining variables $X_4$, $X_5$, $X_6$, $X_7$ are all uniform error, with small variance. `scurve_umap` is the UMAP $2\text{-}D$ embedding for `scurve` data with `n_neighbors` is $46$ and `min_dist` is $0.9$. Each data set contains a unique ID column that maps `scurve` and `scurve_umap`.

@@

Finally, a tibble is returned containing a unique hexagon ID (`h`) along with the corresponding x and y centroid coordinates (`c_x`, `c_y`), which define the layout of the hexagonal grid over the $2-\text{D}$ space.

@@

After generating the centroids that define the hexagonal grid, the next step is to assign each point in the NLDR embedding to its nearest hexagonal bin. The `assign_data()` function performs this assignment by calculating the $2-\text{D}$ Euclidean distance between each point in the $2-\text{D}$ embedding and all hexagon centroids.

First, the function extracts the first two dimensions of the scaled NLDR embedding, which represent the $2-\text{D}$ layout. It then selects the corresponding x and y coordinates of each hexagon’s centroid.

@@

The result is a data frame of the scaled $2-\text{D}$ embedding with an additional `h` column, indicating the hexagonal bin to which each point belongs.

@@

The function first groups the input data by `h`, which represents the hexagon label associated with each point in the $2-\text{D}$ layout. Within each group, it collects the `ID`s into a list, resulting in a summary data frame where each row corresponds to a single hexagon. The resulting column, `pts_list`, contains all point identifiers associated with that hexagon.

@@

To represent the neighborhood structure of hexagonal bins in a $2-\text{D}$ layout, we employ Delaunay triangulation on the centroids of hexagons. This geometric approach is used to infer which bins are considered neighbors.

@@

Not all hexagons contain meaningful information. Some may have very few or no data points due to the sparsity or shape of the underlying structure. Simply removing hexagons with low counts (e.g., fewer than a fixed threshold) can lead to gaps or "holes" in the $2-\text{D}$ structure, potentially disrupting the continuity of the representation.

@@

The function starts by renaming columns in the `model_highd` data frame to avoid conflicts. It then predicts the high-dimensional bin assignments for each point using the `predict_emb()` function, mapping each observation to its nearest $2-\text{D}$ bin (`pred_hb_id`). This prediction is joined with the mean high-dimensional coordinates of each bin from the model and with the original high-dimensional data.

@@

The package offers several $2-\text{D}$ visualisations, including:

@@

The generated p-D model, overlaid with the data, can also be visualized using `show_langevitour`. Additionally, it features a function for visualizing the $2-\text{D}$ projection of the fitted model overlaid on the data, called `plot_proj`.

@@

In the resulting interactive visualisation black points represent the high-dimensional data, green points represent the model centroids from each bin, and the lines between model points reflect the $2-\text{D}$ wireframe structure mapped back to high-dimensional space.

@@

Before visualisation, the input data must be prepared using the `comb_all_data_model()` function. This function combines the high-dimensional data (`highd_data`), the NLDR data (`nldr_data`), and the bin-averaged high-dimensional model representation (`model_highd`) aligned to the $2-\text{D}$ bin layout (`model_2d`):

@@

The function `show_link_plots()` generates two side-by-side, interactively linked plots; a $2-\text{D}$ NLDR representation, and a dynamic projection tour in the original high-dimensional space (using the `langevitour` package), displaying both the data and the model. The function takes the output from `comb_all_data_model()` (`point_data`) and `edge_data` which defines connections between neighboring bins.

@@

Before visualisation, the input data must be prepared using the `comb_all_data_model_error()` function. The function requires several arguments: points data which contain high-dimensional data (`highd_data`), NLDR data (`nldr_data`), high-dimensional model data (`model_highd`), $2-\text{D}$ model data (`model_2d`), and model error (`error_data`).

@@

The function `show_error_link_plots()` generates three side-by-side, interactively linked plots; a error distribution, a $2-\text{D}$ NLDR representation, and a dynamic projection tour in the original high-dimensional space (using the `langevitour` package), displaying both the data and the model. The function takes the output from `comb_all_data_model_error()` (`point_data`) and `edge_data` which defines connections between neighboring bins.

@@

Several core computations within `quollr` are optimized using compiled C++ code via the `Rcpp` and `RcppArmadillo` packages. While the user interacts with high-level R functions, performance-critical steps such as nearest-neighbor searches (`compute_highd_dist()`), error metrics (`compute_errors()`), $2-\text{D}$ distance calculations (`calc_2d_dist_cpp()`), and generation of hexagon coordinates (`gen_hex_coord_cpp()`) are handled internally in C++. This design provides significant speedups when analyzing large datasets while maintaining a user-friendly R interface. These C++ functions are not exported but are bundled within the package and fully accessible for inspection in the source code.

@@

Limb muscle cells of mice in @tabula2018 are examined. There are $1067$ single cells, with $14997$ gene expressions. Following their pre-processing, different NLDR methods were performed using ten principal components. Figure \@ref(fig:limb-rmse) (b) is the reproduction of the published plot. The question is whether this accurately represents the cluster structure in the data. Our method help to provide a better $2-\text{D}$ layout.

@@

The `quollr` package introduces a new framework for interpreting NLDR outputs by fitting a geometric wireframe model in $2-\text{D}$ and lifting it into high-dimensional space. This lifted model provides a direct way to assess how well a $2-\text{D}$ layout, produced by methods such as tSNE, UMAP, PHATE, TriMAP, or PaCMAP, preserves the structure of the original high-dimensional data. The approach offers both numerical and visual diagnostics to support the selection of NLDR methods and tuning hyper-parameters that produce the most accurate $2-\text{D}$ representations.

@@

To support efficient computation, particularly for large-scale datasets, several core functions in `quollr` are implemented in C++ using `Rcpp` and `RcppArmadillo`. These include functions for computing Euclidean distances in high-dimensional and $2-\text{D}$ space, identifying nearest centroids, calculating residual errors, and generating polygonal coordinates of hexagons. For instance, `compute_highd_dist()` accelerates nearest neighbor lookup in high-dimensional space, `compute_errors()` calculates RMSE and total absolute error efficiently, and `calc_2d_dist_cpp()` speeds up distance calculations in $2-\text{D}$. Additionally, `gen_hex_coord_cpp()` constructs the coordinates for hexagonal bins based on their centroids with minimal overhead. These optimizations result in substantial performance gains compared to native R implementations, making the package responsive even when used in interactive contexts or on large datasets such as single-cell transcriptomic profiles.

>>>>

scurve_model_obj <- fit_highd_model(
  highd_data = scurve, 
  nldr_data = scurve_umap, 
  b1 = 15, 
  q = 0.1, 
  benchmark_highdens = 1)

@@

The paper is organized as follows. The next section introduces the implementation of the `quollr` package on CRAN, including a demonstration of the package's key functions and visualization capabilities. In the application section, we illustrate the algorithm's functionality for studying a clustering data structure. Finally, we conclude the paper with a brief summary and discuss potential opportunities for using our algorithm.

@@

Our algorithm includes the following steps: (1) scaling the NLDR data, (2) computing configurations of a hexagon grid, (3) binning the data, (4) obtaining the centroids of each bin, (5) indicating neighboring bins with line segments that connect the centroids, and (6) lifting the model into high dimensions (`r knitr::asis_output(if (knitr::is_html_output()) { "Figure \\@ref(fig:algo-step-html)"} else { "Figure \\@ref(fig:algo-steps)"})`). A detailed description of the algorithm can be found in @gamage2025c.

@@

`scurve` is a $7\text{-}D$ simulated dataset. It is constructed by simulating $1000$ observations from $\theta \sim U(-3\pi/2, 3\pi/2)$, $X_1 = \sin(\theta)$, $X_2 \sim U(0, 2)$ (adding thickness to the S), $X_3 = \text{sign}(\theta) \times (\cos(\theta) - 1)$. The remaining variables $X_4$, $X_5$, $X_6$, $X_7$ are all uniform error, with small variance. `scurve_umap` is the UMAP $2\text{-}D$ embedding for `scurve` data with `n_neighbors` is $46$ and `min_dist` is $0.9$. Each data set contains a unique ID column that maps `scurve` and `scurve_umap`.

@@

Finally, a tibble is returned containing a unique hexagon ID (`h`) along with the corresponding x and y centroid coordinates (`c_x`, `c_y`), which define the layout of the hexagonal grid over the $2\text{-}D$ space.

@@

After generating the centroids that define the hexagonal grid, the next step is to assign each point in the NLDR embedding to its nearest hexagonal bin. The `assign_data()` function performs this assignment by calculating the $2\text{-}D$ Euclidean distance between each point in the $2\text{-}D$ embedding and all hexagon centroids.

First, the function extracts the first two dimensions of the scaled NLDR embedding, which represent the $2\text{-}D$ layout. It then selects the corresponding x and y coordinates of each hexagon’s centroid.

@@

The result is a data frame of the scaled $2\text{-}D$ embedding with an additional `h` column, indicating the hexagonal bin to which each point belongs.

@@

The function first groups the input data by `h`, which represents the hexagon label associated with each point in the $2\text{-}D$ layout. Within each group, it collects the `ID`s into a list, resulting in a summary data frame where each row corresponds to a single hexagon. The resulting column, `pts_list`, contains all point identifiers associated with that hexagon.

@@

To represent the neighborhood structure of hexagonal bins in a $2\text{-}D$ layout, we employ Delaunay triangulation on the centroids of hexagons. This geometric approach is used to infer which bins are considered neighbors.

@@

Not all hexagons contain meaningful information. Some may have very few or no data points due to the sparsity or shape of the underlying structure. Simply removing hexagons with low counts (e.g., fewer than a fixed threshold) can lead to gaps or "holes" in the $2\text{-}D$ structure, potentially disrupting the continuity of the representation.

@@

The function starts by renaming columns in the `model_highd` data frame to avoid conflicts. It then predicts the high-dimensional bin assignments for each point using the `predict_emb()` function, mapping each observation to its nearest $2\text{-}D$ bin (`pred_hb_id`). This prediction is joined with the mean high-dimensional coordinates of each bin from the model and with the original high-dimensional data.

@@

The package offers several $2\text{-}D$ visualisations, including:

@@

The generated p-D model, overlaid with the data, can also be visualized using `show_langevitour`. Additionally, it features a function for visualizing the $2\text{-}D$ projection of the fitted model overlaid on the data, called `plot_proj`.

@@

In the resulting interactive visualisation black points represent the high-dimensional data, green points represent the model centroids from each bin, and the lines between model points reflect the $2\text{-}D$ wireframe structure mapped back to high-dimensional space.

@@

Before visualisation, the input data must be prepared using the `comb_all_data_model()` function. This function combines the high-dimensional data (`highd_data`), the NLDR data (`nldr_data`), and the bin-averaged high-dimensional model representation (`model_highd`) aligned to the $2\text{-}D$ bin layout (`model_2d`):

@@

The function `show_link_plots()` generates two side-by-side, interactively linked plots; a $2\text{-}D$ NLDR representation, and a dynamic projection tour in the original high-dimensional space (using the `langevitour` package), displaying both the data and the model. The function takes the output from `comb_all_data_model()` (`point_data`) and `edge_data` which defines connections between neighboring bins.

@@

Before visualisation, the input data must be prepared using the `comb_all_data_model_error()` function. The function requires several arguments: points data which contain high-dimensional data (`highd_data`), NLDR data (`nldr_data`), high-dimensional model data (`model_highd`), $2\text{-}D$ model data (`model_2d`), and model error (`error_data`).

@@

The function `show_error_link_plots()` generates three side-by-side, interactively linked plots; a error distribution, a $2\text{-}D$ NLDR representation, and a dynamic projection tour in the original high-dimensional space (using the `langevitour` package), displaying both the data and the model. The function takes the output from `comb_all_data_model_error()` (`point_data`) and `edge_data` which defines connections between neighboring bins.

@@

Several core computations within `quollr` are optimized using compiled C++ code via the `Rcpp` and `RcppArmadillo` packages. While the user interacts with high-level R functions, performance-critical steps such as nearest-neighbor searches (`compute_highd_dist()`), error metrics (`compute_errors()`), $2\text{-}D$ distance calculations (`calc_2d_dist_cpp()`), and generation of hexagon coordinates (`gen_hex_coord_cpp()`) are handled internally in C++. This design provides significant speedups when analyzing large datasets while maintaining a user-friendly R interface. These C++ functions are not exported but are bundled within the package and fully accessible for inspection in the source code.

@@

Limb muscle cells of mice in @tabula2018 are examined. There are $1067$ single cells, with $14997$ gene expressions. Following their pre-processing, different NLDR methods were performed using ten principal components. Figure \@ref(fig:limb-rmse) (b) is the reproduction of the published plot. The question is whether this accurately represents the cluster structure in the data. Our method help to provide a better $2\text{-}D$ layout.

@@

The `quollr` package introduces a new framework for interpreting NLDR outputs by fitting a geometric wireframe model in $2\text{-}D$ and lifting it into high-dimensional space. This lifted model provides a direct way to assess how well a $2\text{-}D$ layout, produced by methods such as tSNE, UMAP, PHATE, TriMAP, or PaCMAP, preserves the structure of the original high-dimensional data. The approach offers both numerical and visual diagnostics to support the selection of NLDR methods and tuning hyper-parameters that produce the most accurate $2\text{-}D$ representations.

@@

To support efficient computation, particularly for large-scale datasets, several core functions in `quollr` are implemented in C++ using `Rcpp` and `RcppArmadillo`. These include functions for computing Euclidean distances in high-dimensional and $2\text{-}D$ space, identifying nearest centroids, calculating residual errors, and generating polygonal coordinates of hexagons. For instance, `compute_highd_dist()` accelerates nearest neighbor lookup in high-dimensional space, `compute_errors()` calculates RMSE and total absolute error efficiently, and `calc_2d_dist_cpp()` speeds up distance calculations in $2\text{-}D$. Additionally, `gen_hex_coord_cpp()` constructs the coordinates for hexagonal bins based on their centroids with minimal overhead. These optimizations result in substantial performance gains compared to native R implementations, making the package responsive even when used in interactive contexts or on large datasets such as single-cell transcriptomic profiles.

### 2025-08-04 paper-quollr.tex
Fixed some issue with new data

The paper is organized as follows. The next section introduces the implementation of the \texttt{quollr} package on CRAN, including a demonstration of the package's key functions and visualisation capabilities. In the application section, we illustrate the algorithm's functionality for studying a clustering data structure. Finally, we conclude the paper with a brief summary and discuss potential opportunities for using our algorithm.

@@

Our algorithm includes the following steps: (1) scaling the NLDR data, (2) computing configurations of a hexagon grid, (3) binning the data, (4) obtaining the centroids of each bin, (5) indicating neighboring bins with line segments that connect the centroids, and (6) lifting the model into high dimensions (Figure \ref{fig:algo-steps}). A detailed description of the algorithm can be found in \textbf{cite the methodology paper}.

@@

\texttt{scurve} is a \(7\text{-}D\) simulated dataset. It is constructed by simulating \(5000\) observations from \(\theta \sim U(-3\pi/2, 3\pi/2)\), \(X_1 = \sin(\theta)\), \(X_2 \sim U(0, 2)\) (adding thickness to the S), \(X_3 = \text{sign}(\theta) \times (\cos(\theta) - 1)\). The remaining variables \(X_4\), \(X_5\), \(X_6\), \(X_7\) are all uniform error, with small variance. \texttt{scurve\_umap} is the UMAP \(2\text{-}D\) embedding for \texttt{scurve} data with \texttt{n\_neighbors} is \(46\) and \texttt{min\_dist} is \(0.9\). Each data set contains a unique ID column that maps \texttt{scurve} and \texttt{scurve\_umap}.

@@

Finally, a tibble is returned containing a unique hexagon ID (\texttt{h}) along with the corresponding x and y centroid coordinates (\texttt{c\_x}, \texttt{c\_y}), which define the layout of the hexagonal grid over the \(2-\text{D}\) space.

@@

After generating the centroids that define the hexagonal grid, the next step is to assign each point in the NLDR embedding to its nearest hexagonal bin. The \texttt{assign\_data()} function performs this assignment by calculating the \(2-\text{D}\) Euclidean distance between each point in the \(2-\text{D}\) embedding and all hexagon centroids.

First, the function extracts the first two dimensions of the scaled NLDR embedding, which represent the \(2-\text{D}\) layout. It then selects the corresponding x and y coordinates of each hexagon's centroid.

@@

The result is a data frame of the scaled \(2-\text{D}\) embedding with an additional \texttt{h} column, indicating the hexagonal bin to which each point belongs.

@@

The function first groups the input data by \texttt{h}, which represents the hexagon label associated with each point in the \(2-\text{D}\) layout. Within each group, it collects the \texttt{ID}s into a list, resulting in a summary data frame where each row corresponds to a single hexagon. The resulting column, \texttt{pts\_list}, contains all point identifiers associated with that hexagon.

@@

To represent the neighborhood structure of hexagonal bins in a \(2-\text{D}\) layout, we employ Delaunay triangulation on the centroids of hexagons. This geometric approach is used to infer which bins are considered neighbors.

@@

Not all hexagons contain meaningful information. Some may have very few or no data points due to the sparsity or shape of the underlying structure. Simply removing hexagons with low counts (e.g., fewer than a fixed threshold) can lead to gaps or ``holes'' in the \(2-\text{D}\) structure, potentially disrupting the continuity of the representation.

@@

The function starts by renaming columns in the \texttt{model\_highd} data frame to avoid conflicts. It then predicts the high-dimensional bin assignments for each point using the \texttt{predict\_emb()} function, mapping each observation to its nearest \(2-\text{D}\) bin (\texttt{pred\_hb\_id}). This prediction is joined with the mean high-dimensional coordinates of each bin from the model and with the original high-dimensional data.

@@

The package offers several \(2-\text{D}\) visualisations, including:

@@

The generated p-D model, overlaid with the data, can also be visualized using \texttt{show\_langevitour}. Additionally, it features a function for visualizing the \(2-\text{D}\) projection of the fitted model overlaid on the data, called \texttt{plot\_proj}.

@@

In the resulting interactive visualisation black points represent the high-dimensional data, green points represent the model centroids from each bin, and the lines between model points reflect the \(2-\text{D}\) wireframe structure mapped back to high-dimensional space.

@@

Before visualisation, the input data must be prepared using the \texttt{comb\_all\_data\_model()} function. This function combines the high-dimensional data (\texttt{highd\_data}), the NLDR data (\texttt{nldr\_data}), and the bin-averaged high-dimensional model representation (\texttt{model\_highd}) aligned to the \(2-\text{D}\) bin layout (\texttt{model\_2d}):

@@

The function \texttt{show\_link\_plots()} generates two side-by-side, interactively linked plots; a \(2-\text{D}\) NLDR representation, and a dynamic projection tour in the original high-dimensional space (using the \texttt{langevitour} package), displaying both the data and the model. The function takes the output from \texttt{comb\_all\_data\_model()} (\texttt{point\_data}) and \texttt{edge\_data} which defines connections between neighboring bins.

@@

Before visualisation, the input data must be prepared using the \texttt{comb\_all\_data\_model\_error()} function. The function requires several arguments: points data which contain high-dimensional data (\texttt{highd\_data}), NLDR data (\texttt{nldr\_data}), high-dimensional model data (\texttt{model\_highd}), \(2-\text{D}\) model data (\texttt{model\_2d}), and model error (\texttt{error\_data}).

@@

The function \texttt{show\_error\_link\_plots()} generates three side-by-side, interactively linked plots; a error distribution, a \(2-\text{D}\) NLDR representation, and a dynamic projection tour in the original high-dimensional space (using the \texttt{langevitour} package), displaying both the data and the model. The function takes the output from \texttt{comb\_all\_data\_model\_error()} (\texttt{point\_data}) and \texttt{edge\_data} which defines connections between neighboring bins.

@@

Several core computations within \texttt{quollr} are optimized using compiled C++ code via the \texttt{Rcpp} and \texttt{RcppArmadillo} packages. While the user interacts with high-level R functions, performance-critical steps such as nearest-neighbor searches (\texttt{compute\_highd\_dist()}), error metrics (\texttt{compute\_errors()}), \(2-\text{D}\) distance calculations (\texttt{calc\_2d\_dist\_cpp()}), and generation of hexagon coordinates (\texttt{gen\_hex\_coord\_cpp()}) are handled internally in C++. This design provides significant speedups when analyzing large datasets while maintaining a user-friendly R interface. These C++ functions are not exported but are bundled within the package and fully accessible for inspection in the source code.

@@

Limb muscle cells of mice in \citet{tabula2018} are examined. There are \(1067\) single cells, with \(14997\) gene expressions. Following their pre-processing, different NLDR methods were performed using ten principal components. Figure \ref{fig:limb-rmse} (b) is the reproduction of the published plot. The question is whether this accurately represents the cluster structure in the data. Our method help to provide a better \(2-\text{D}\) layout.

@@

The \texttt{quollr} package introduces a new framework for interpreting NLDR outputs by fitting a geometric wireframe model in \(2-\text{D}\) and lifting it into high-dimensional space. This lifted model provides a direct way to assess how well a \(2-\text{D}\) layout, produced by methods such as tSNE, UMAP, PHATE, TriMAP, or PaCMAP, preserves the structure of the original high-dimensional data. The approach offers both numerical and visual diagnostics to support the selection of NLDR methods and tuning hyper-parameters that produce the most accurate \(2-\text{D}\) representations.

@@

To support efficient computation, particularly for large-scale datasets, several core functions in \texttt{quollr} are implemented in C++ using \texttt{Rcpp} and \texttt{RcppArmadillo}. These include functions for computing Euclidean distances in high-dimensional and \(2-\text{D}\) space, identifying nearest centroids, calculating residual errors, and generating polygonal coordinates of hexagons. For instance, \texttt{compute\_highd\_dist()} accelerates nearest neighbor lookup in high-dimensional space, \texttt{compute\_errors()} calculates RMSE and total absolute error efficiently, and \texttt{calc\_2d\_dist\_cpp()} speeds up distance calculations in \(2-\text{D}\). Additionally, \texttt{gen\_hex\_coord\_cpp()} constructs the coordinates for hexagonal bins based on their centroids with minimal overhead. These optimizations result in substantial performance gains compared to native R implementations, making the package responsive even when used in interactive contexts or on large datasets such as single-cell transcriptomic profiles.

>>>>

The paper is organized as follows. The next section introduces the implementation of the \texttt{quollr} package on CRAN, including a demonstration of the package's key functions and visualization capabilities. In the application section, we illustrate the algorithm's functionality for studying a clustering data structure. Finally, we conclude the paper with a brief summary and discuss potential opportunities for using our algorithm.

@@

Our algorithm includes the following steps: (1) scaling the NLDR data, (2) computing configurations of a hexagon grid, (3) binning the data, (4) obtaining the centroids of each bin, (5) indicating neighboring bins with line segments that connect the centroids, and (6) lifting the model into high dimensions (Figure \ref{fig:algo-steps}). A detailed description of the algorithm can be found in \citet{gamage2025c}.

@@

\texttt{scurve} is a \(7\text{-}D\) simulated dataset. It is constructed by simulating \(1000\) observations from \(\theta \sim U(-3\pi/2, 3\pi/2)\), \(X_1 = \sin(\theta)\), \(X_2 \sim U(0, 2)\) (adding thickness to the S), \(X_3 = \text{sign}(\theta) \times (\cos(\theta) - 1)\). The remaining variables \(X_4\), \(X_5\), \(X_6\), \(X_7\) are all uniform error, with small variance. \texttt{scurve\_umap} is the UMAP \(2\text{-}D\) embedding for \texttt{scurve} data with \texttt{n\_neighbors} is \(46\) and \texttt{min\_dist} is \(0.9\). Each data set contains a unique ID column that maps \texttt{scurve} and \texttt{scurve\_umap}.

@@

Finally, a tibble is returned containing a unique hexagon ID (\texttt{h}) along with the corresponding x and y centroid coordinates (\texttt{c\_x}, \texttt{c\_y}), which define the layout of the hexagonal grid over the \(2\text{-}D\) space.

@@

After generating the centroids that define the hexagonal grid, the next step is to assign each point in the NLDR embedding to its nearest hexagonal bin. The \texttt{assign\_data()} function performs this assignment by calculating the \(2\text{-}D\) Euclidean distance between each point in the \(2\text{-}D\) embedding and all hexagon centroids.

First, the function extracts the first two dimensions of the scaled NLDR embedding, which represent the \(2\text{-}D\) layout. It then selects the corresponding x and y coordinates of each hexagon's centroid.

@@

The result is a data frame of the scaled \(2\text{-}D\) embedding with an additional \texttt{h} column, indicating the hexagonal bin to which each point belongs.

@@

The function first groups the input data by \texttt{h}, which represents the hexagon label associated with each point in the \(2\text{-}D\) layout. Within each group, it collects the \texttt{ID}s into a list, resulting in a summary data frame where each row corresponds to a single hexagon. The resulting column, \texttt{pts\_list}, contains all point identifiers associated with that hexagon.

@@

To represent the neighborhood structure of hexagonal bins in a \(2\text{-}D\) layout, we employ Delaunay triangulation on the centroids of hexagons. This geometric approach is used to infer which bins are considered neighbors.

@@

Not all hexagons contain meaningful information. Some may have very few or no data points due to the sparsity or shape of the underlying structure. Simply removing hexagons with low counts (e.g., fewer than a fixed threshold) can lead to gaps or ``holes'' in the \(2\text{-}D\) structure, potentially disrupting the continuity of the representation.

@@

The function starts by renaming columns in the \texttt{model\_highd} data frame to avoid conflicts. It then predicts the high-dimensional bin assignments for each point using the \texttt{predict\_emb()} function, mapping each observation to its nearest \(2\text{-}D\) bin (\texttt{pred\_hb\_id}). This prediction is joined with the mean high-dimensional coordinates of each bin from the model and with the original high-dimensional data.

@@

The package offers several \(2\text{-}D\) visualisations, including:

@@

The generated p-D model, overlaid with the data, can also be visualized using \texttt{show\_langevitour}. Additionally, it features a function for visualizing the \(2\text{-}D\) projection of the fitted model overlaid on the data, called \texttt{plot\_proj}.

@@

In the resulting interactive visualisation black points represent the high-dimensional data, green points represent the model centroids from each bin, and the lines between model points reflect the \(2\text{-}D\) wireframe structure mapped back to high-dimensional space.

@@

Before visualisation, the input data must be prepared using the \texttt{comb\_all\_data\_model()} function. This function combines the high-dimensional data (\texttt{highd\_data}), the NLDR data (\texttt{nldr\_data}), and the bin-averaged high-dimensional model representation (\texttt{model\_highd}) aligned to the \(2\text{-}D\) bin layout (\texttt{model\_2d}):

@@

The function \texttt{show\_link\_plots()} generates two side-by-side, interactively linked plots; a \(2\text{-}D\) NLDR representation, and a dynamic projection tour in the original high-dimensional space (using the \texttt{langevitour} package), displaying both the data and the model. The function takes the output from \texttt{comb\_all\_data\_model()} (\texttt{point\_data}) and \texttt{edge\_data} which defines connections between neighboring bins.

@@

Before visualisation, the input data must be prepared using the \texttt{comb\_all\_data\_model\_error()} function. The function requires several arguments: points data which contain high-dimensional data (\texttt{highd\_data}), NLDR data (\texttt{nldr\_data}), high-dimensional model data (\texttt{model\_highd}), \(2\text{-}D\) model data (\texttt{model\_2d}), and model error (\texttt{error\_data}).

@@

The function \texttt{show\_error\_link\_plots()} generates three side-by-side, interactively linked plots; a error distribution, a \(2\text{-}D\) NLDR representation, and a dynamic projection tour in the original high-dimensional space (using the \texttt{langevitour} package), displaying both the data and the model. The function takes the output from \texttt{comb\_all\_data\_model\_error()} (\texttt{point\_data}) and \texttt{edge\_data} which defines connections between neighboring bins.

@@

Several core computations within \texttt{quollr} are optimized using compiled C++ code via the \texttt{Rcpp} and \texttt{RcppArmadillo} packages. While the user interacts with high-level R functions, performance-critical steps such as nearest-neighbor searches (\texttt{compute\_highd\_dist()}), error metrics (\texttt{compute\_errors()}), \(2\text{-}D\) distance calculations (\texttt{calc\_2d\_dist\_cpp()}), and generation of hexagon coordinates (\texttt{gen\_hex\_coord\_cpp()}) are handled internally in C++. This design provides significant speedups when analyzing large datasets while maintaining a user-friendly R interface. These C++ functions are not exported but are bundled within the package and fully accessible for inspection in the source code.

@@

Limb muscle cells of mice in \citet{tabula2018} are examined. There are \(1067\) single cells, with \(14997\) gene expressions. Following their pre-processing, different NLDR methods were performed using ten principal components. Figure \ref{fig:limb-rmse} (b) is the reproduction of the published plot. The question is whether this accurately represents the cluster structure in the data. Our method help to provide a better \(2\text{-}D\) layout.

@@

The \texttt{quollr} package introduces a new framework for interpreting NLDR outputs by fitting a geometric wireframe model in \(2\text{-}D\) and lifting it into high-dimensional space. This lifted model provides a direct way to assess how well a \(2\text{-}D\) layout, produced by methods such as tSNE, UMAP, PHATE, TriMAP, or PaCMAP, preserves the structure of the original high-dimensional data. The approach offers both numerical and visual diagnostics to support the selection of NLDR methods and tuning hyper-parameters that produce the most accurate \(2\text{-}D\) representations.

@@

To support efficient computation, particularly for large-scale datasets, several core functions in \texttt{quollr} are implemented in C++ using \texttt{Rcpp} and \texttt{RcppArmadillo}. These include functions for computing Euclidean distances in high-dimensional and \(2\text{-}D\) space, identifying nearest centroids, calculating residual errors, and generating polygonal coordinates of hexagons. For instance, \texttt{compute\_highd\_dist()} accelerates nearest neighbor lookup in high-dimensional space, \texttt{compute\_errors()} calculates RMSE and total absolute error efficiently, and \texttt{calc\_2d\_dist\_cpp()} speeds up distance calculations in \(2\text{-}D\). Additionally, \texttt{gen\_hex\_coord\_cpp()} constructs the coordinates for hexagonal bins based on their centroids with minimal overhead. These optimizations result in substantial performance gains compared to native R implementations, making the package responsive even when used in interactive contexts or on large datasets such as single-cell transcriptomic profiles.
